Estimating demographic bias on tests of children's early vocabulary

### 2022-09-13 paper/manuscript.Rmd
Mike's edits incorporated

author: 
  - name          : "George Kachergis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "450 Jane Stanford Way"
    email         : "gkacherg@stanford.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Supervision"
  - name          : "Nathan Francis"
    affiliation   : "1"
    role:
      - "Writing - Review & Editing"
  - name          : "Michael C. Frank"
    affiliation   : "1"
#    role:
#      - "Conceptualization"
#      - "Writing - Original Draft Preparation"
#      - "Writing - Review & Editing"
#      - "Supervision"

@@

keywords          : "language acquisition; word learning; measuring instrument bias; demographic bias"
wordcount         : "4008"

@@

One challenge in comprehensive assessment of these differences is that it is difficult to get a complete measure of young children's language skill. Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

@@

<!--Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2021cat].
These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children being reported on) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
The efficacy of these IRT-based models depends on words varying in difficulty, in order for the test to be more informative of the ability level of each individual.
For example, asking whether a 22-month-old produces the word "ball" is far less informative of that child's language ability than asking whether they produce "table", as 96% of 22-month-olds can produce the former, while only 47% produce the latter. -->

@@

The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021]. And both observations could be true -- a test could be biased to inflate a true demographic difference.

Our goal in the current study is to use DIF analyses to assess these possibilities. 
Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs. female), maternal education (no more than secondary vs. at least some college), and race (white vs. non-white).

The outline of this paper is as follows.
First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
We then fit the IRT model to both groups for each demographic factor, and look at the item parameters for evidence of DIF, noting in particular how many items are significantly biased in favor of each group, and describing them qualitatively.
Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI showing demographic bias.

@@

The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model, and is thus the easiest to use to investigate potential differences in item function across different groups of participants.
The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of the sex GLIMMER (Fig. 2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

@@

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in  demographic studies [@rowe2009differences; @huttenlocher1991early].
In an attempt to investigate this, we analyzed the extent to which word frequency in a corpus of child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5213750 female-directed tokens and 6091950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.

>>>>

author: 
  - name          : "George Kachergis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "450 Jane Stanford Way"
    email         : "gkacherg@stanford.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Supervision"
  - name          : "Nathan Francis"
    affiliation   : "1"
    role:
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Michael C. Frank"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Supervision"

@@

keywords          : "language acquisition; word learning; measuring instrument bias; demographic bias"
wordcount         : "3988"

@@

One challenge in comprehensive assessment of these differences is that it is difficult to get a complete measure of young children's language skill. Long recordings are prohibitively difficult to collect and transcribe, yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

@@

<!--Inspired by the utility and widespread use of the CDI, researchers have recently been using psychometric models on CDI data to construct short, adaptive tests to reliably assess language ability using only a small subset of the CDI items [e.g., @mayor2019; @kachergis2022cat].
These psychometric models typically come from the Item-Response Theory (IRT) framework [@Baker2001], which assumes that not only test-takers (here, children being reported on) have normally-distributed ability, but that items (words on the CDI) have normally-distributed difficulty.
The efficacy of these IRT-based models depends on words varying in difficulty, in order for the test to be more informative of the ability level of each individual.
For example, asking whether a 22-month-old produces the word "ball" is far less informative of that child's language ability than asking whether they produce "table", as 96% of 22-month-olds can produce the former, while only 47% produce the latter. -->

@@

The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF) [@holland1993differential]. 
DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
Of course, there may be a true ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021]. 
And both observations could be true -- a test could be biased to inflate *or* deflate a true demographic difference.

Our goal in the current study is to use DIF analyses to assess these possibilities. 
We focus on the American English data from Wordbank due to its large size and due to the presence of demographic data for our variables of interest.
Specifically, we test the 680 vocabulary items on the American English CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs. female), maternal education (no more than secondary vs. at least some college), and race (white vs. non-white).

The outline of this paper is as follows.
First, we introduce the Wordbank data and the IRT model, and examine the overall size of demographic differences in early word learning based on the full CDI:WS.
We then fit the IRT model to both groups for each demographic factor and examine the item parameters for evidence of DIF.
Next, we use these item parameters to systematically prune biased items from the CDI:WS, and examine how the estimated size of demographic differences change as we prune more items.
Finally, using a corpus of child-directed speech, we examine the extent to which variation in word frequencies heard by boys vs. girls predicts sex-based DIF.
Based on these analyses, we provide recommendations for next steps to be taken to identify and potentially replace items on the CDI that show demographic bias.

@@

The Rasch model, also known as the 1-parameter logistic (1PL) model, is the simplest Item Response Theory model.
Although we have elsewhere shown that the 2-parameter logistic model often fits CDI data better [@kachergis2022cat], the simpler Rasch model is easier to use to evaluate potential differences in item function across different groups of participants.
The Rasch model jointly estimates for each child $j$ a latent ability $\theta_j$, and for each item $i$ a difficulty parameter $b_i$.
In the model, the probability of child $j$ knowing (i.e., producing or understanding) a given item $i$ is

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are large but can be found at OSF^[1], but were too large to include here.
Inspecting GLIMMER plots can reveal clusters of items, indicating candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of the sex GLIMMER (Figure 2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

^[1]: [https://osf.io/57rsw/](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a)

@@

Demographic differences in language ability are likely to be at least partially explained by differences in linguistic input received by children in different groups.
Indeed, input quantity (total daily tokens) and some measures of quality (e.g., lexical diversity: ratio of word types vs. tokens) have often been predictive of language learning outcomes in demographic studies [@rowe2009differences; @huttenlocher1991early].
In an attempt to investigate this, we analyzed the extent to which word frequency in a corpus of child-directed speech to male vs. female children was predictive of the amount of sex-based bias shown by CDI items. 
Similar to the approach taken by @braginsky2016gender, we used the CHILDES corpus of transcripts from dyadic play sessions [@macwhinney2000childes], which are labeled with the sex of the target child, but not other demographic variables.
We found a total of 5,213,750 female-directed tokens and 6,091,950 male-directed tokens that matched 662 of the 680 CDI:WS items, normalized the word frequencies to tokens per million for each target sex, and calculated the log-odds of each word being spoken to a female (i.e., $log(p_{f} / p_{m}$), meaning unbiased words will have log-odds of 0, while those spoken more often to females will have log-odds $>0$, and those spoken more often males will have log-odds $<0$.
For example "doll" was spoken 156 times to females, and 138 times to males, thus $p_{f} = 156/(156+138) = .53$, $p_{m} = 1-p_{f} = 0.47$, and $log(p_{f}/p_{m})$ = 0.12), while "police" was spoken 138 times to females, and 225 times to males, and thus has log-odds = -0.49).
Overall, the correlation between the log-odds of a word being spoken to a female vs. a male child and the size of the female (vs. male) advantage for that CDI word was weak, but significant ($r=0.18$, $p<.001$). 
Some of the sex bias seen in CDI items is related to differences in language input received by girls vs. boys -- though of course this differential input could be elicited due to different interests.
